The dynamics of starvation and recovery

### 2017-06-10 Manuscript/Nature-Submission/Nature-temp.tex
nature comm

\begin{abstract}
The eco-evolutionary dynamics of species are fundamentally linked to the energetic constraints of its constituent individuals. Of particular importance are the tradeoffs between reproduction and the dynamics of starvation and recovery in resource-limited environments. To elucidate the consequences of this tradeoff, we introduce a minimal nutritional state-structured model that incorporates two classes of consumer: nutritionally replete consumers that reproduce, and undernourished, non-reproducing consumers that are susceptible to mortality. As a function of the transition rates between these two states that are determined by the abundance of resources, the consumer populations can either undergo cyclic dynamics or reach a steady state. We obtain strong constraints on starvation and recovery rates by deriving allometric scaling relationships between body size and a variety of traits and find that population dynamics subject to these constraints are typically driven to a steady state. Moreover, we find that these rates fall within a `refuge' in parameter space, where the probability of extinction of the consumer population is minimized. Thus we identify a potential mechanism that may both drive and constrain the dynamics of animal populations. Our model provides a natural framework that predicts maximum body size for mammals by determining the relative stability of an otherwise homogeneous population to a mutant population with altered percentage of body fat. For body masses $< 1.748\times10^7$g, individuals with increased energetic reserves can invade resident populations, and vice versa for body mass $> 1.748\times10^7$g, thus providing a principled mechanism for a within-lineage driver of Cope's rule.
\end{abstract}

%\noindent {\small {\bf Significance Statement} %105/120 words
%Energetic investment in somatic maintenance and growth vs. reproduction directly impacts the dynamics of populations among species. Here, we construct a Nutritional State-structured Model (NSM) to assess the population-level effects of starvation and recovery of a consumer population in a resource-limited environment, and use allometric scaling relationships for mammals to establish all timescales and rates. Our model: i. reveals that mammalian energetic rates minimize the probability of stochastic extinction, ii. establishes dynamic bounds on mammalian body size while providing independent theoretical support for the energy equivalence hypothesis, and iii. provides a mechanistic driver for the evolutionary trend towards larger body size known as Cope's rule.}

\noindent {\bf Introduction} \\
The behavioral ecology of all organisms is influenced by the energetic state of individuals, which directly influences how they invest reserves in uncertain environments.
Such behaviors are generally manifested as tradeoffs between investing in somatic maintenance and growth, or allocating energy towards reproduction~\cite{Martin:1987dl,Kirk:1997cc,Kempes:2012hy}.
The timing of these behaviors responds to selective pressure, as the choice of the investment impacts future fitness~\cite{Mangel:1988uaa,Mangel:2014kz,Yeakel:2013hi}.
The influence of resource limitation on an organism's ability to maintain its nutritional stores may lead to repeated delays or shifts in reproduction over the course of an organism's life.

>>>>

\begin{abstract}
The behavioral ecology of all organisms is influenced by the energetic state of individuals, which directly influences how they invest reserves in uncertain environments.
Such behaviors are generally manifested as a tradeoff between investing in somatic maintenance and growth, or allocating energy towards reproduction~\cite{Martin:1987dl,Kirk:1997cc,Kempes:2012hy}.
% Of particular importance are the tradeoffs between reproduction and the dynamics of starvation and recovery in resource-limited environments. 
To elucidate the consequences of this tradeoff, we introduce a minimal nutritional state-structured model that incorporates two classes of consumer: nutritionally replete consumers that reproduce, and undernourished, non-reproducing consumers that are susceptible to mortality. 
% As a function of the transition rates between these two states that are determined by the abundance of resources, the consumer populations can either undergo cyclic dynamics or reach a steady state. 
We obtain strong constraints on starvation and recovery rates by deriving allometric scaling relationships between body size and a variety of traits and find that population dynamics subject to these constraints are typically driven to a steady state. 
Moreover, we find that these rates fall within a `refuge' in parameter space, where the probability of extinction of the consumer population is minimized. 
% Thus we identify a potential mechanism that may both drive and constrain the dynamics of animal populations. 
Our model provides a natural framework that predicts maximum body size for mammals by determining the relative stability of an otherwise homogeneous population to a mutant population with altered percentage of body fat. 
% For body masses $< 1.748\times10^7$g, individuals with increased energetic reserves can invade resident populations, and vice versa for body mass $> 1.748\times10^7$g, thus providing a principled mechanism for a within-lineage driver of Cope's rule.
\end{abstract}

% \noindent {\small {\bf Summary} %105/120 words
%Energetic investment in somatic maintenance and growth vs. reproduction directly impacts the dynamics of populations among species. Here, we construct a Nutritional State-structured Model (NSM) to assess the population-level effects of starvation and recovery of a consumer population in a resource-limited environment, and use allometric scaling relationships for mammals to establish all timescales and rates. Our model: i. reveals that mammalian energetic rates minimize the probability of stochastic extinction, ii. establishes dynamic bounds on mammalian body size while providing independent theoretical support for the energy equivalence hypothesis, and iii. provides a mechanistic driver for the evolutionary trend towards larger body size known as Cope's rule.}

% \noindent {\bf Introduction} \\

% The timing of these behaviors responds to selective pressure, as the choice of the investment impacts future fitness~\cite{Mangel:1988uaa,Mangel:2014kz,Yeakel:2013hi}.
% The influence of resource limitation on an organism's ability to maintain its nutritional stores may lead to repeated delays or shifts in reproduction over the course of an organism's life.
